---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Graham, Hannah"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

#str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME
str(mydata)
```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
#Use *summary()* to obtain and present descriptive statistics from mydata.
summary(mydata)

#Use *table()* to present a frequency table using CLASS and RINGS
Class <- mydata$CLASS
Rings <- mydata$RINGS

tableClassVRings <- table(Class, Rings)
tableClassVRings

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Abalones are a type of snail. Whole pertains to the weight of the abalone in grams, shuck is the weight of the meat, Rings pertains to age of abalones that are over 1.5 years. Class is a classification of age, based on the rings. Class has a very broad range for A5, where over 13 different ring values ar included in this class, as compared to only 2-4 ring values for the classes a1=a4. A5, I predict, will hold many of the outliers. The majority of the abalones belong in the A3 class out of all the classes, at 9 and 10 rings.  ***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
#Generate a table of counts using SEX and CLASS
Sex <- mydata$SEX
Class <- mydata$CLASS
sexVclassTable <- table(Sex, Class)

#Add margins to this table
margins <- addmargins(sexVclassTable)
sexVclassTable

#Lastly, present a barplot of these data; ignoring the marginal totals.
my_bar <- barplot(sexVclassTable,
                ylab = "Frequency",
                xlab = "Class",
                main = "Frequency of Sex by Class",
                 beside = TRUE,
                 legend.text = TRUE,
                 cex.names = TRUE,
                 col = c("black", "mistyrose", "red")
                  )


```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (Infant abalones class is inversely proportionate to Female and Male abalone classes. Infants have a higher frequency in A1&A2 and decreases on, while Female and male have a sort of bell curve style where the majority of them lie in the class of A 3. Its also interesting to me that A4 and A5 class for female and male don't have a substantial decrease in frequency, meaning if an abalone makes it to class A4 in age, it will probably make it to A5. It seems from A1 to A3 there is a significant more amount of female abalones, but then that levels out in A4 and A5 so that there is almost an equal frequency of female and male abalones present. )***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
#Use *set.seed(123)* prior to drawing this sample
set.seed(123)
#We need to sample from the integers, 1 to 1036, representing the rows of "mydata.
BigRange <- seq(1:1036)

#elect a simple random sample of 200 observations from "mydata" and identify this sample as "work." 
sampleInd <- sample(BigRange, 200)
work <- mydata[sampleInd,]

#sing "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])*
plot(work[, 2:6])



```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
#Use "mydata" to plot WHOLE versus VOLUME
plot(mydata$WHOLE,
     mydata$VOLUME,
     col= mydata$CLASS,
     xlab= "Whole Weight",
     ylab= "Volumne",
     main = "Whole Weight by Volume",
     cex = 0.5
     )
#add legend for Class.
legend("topright", legend=levels(mydata$CLASS), col=1:length(levels(mydata$CLASS)), pch=1)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
# Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS
plot(
  mydata$WHOLE,
  mydata$SHUCK,
  col=mydata$CLASS,
  cex=0.5,
  xlab = "Whole Weight",
  ylab = "Shuck Weight",
  main = "Shuck Weight by Whole Weight",
)
#As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE
maxRatio <- max(mydata$SHUCK/mydata$WHOLE)
maxRatio

#Add to the chart a straight line with zero intercept using this maximum value as the slope of the line.
abline(0, maxRatio)

#Ensure legend for Class is present
legend("topright", legend=levels(mydata$CLASS), col=1:length(levels(mydata$CLASS)), pch=1)
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (In part A, it seems whole weight compared to volume seems quite evenly mixed. Where in part B, Shuck Weight compared to whole weight has clear correlation with Class. For example, A2, closely follows the mas ration line, then just below it is A3, then A4. As whole weight increases, Shuck weight diverges further and further from the max ratio line. )***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
#Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex

#histograms
histogramSex <- ggplot(mydata, aes(RATIO, fill=SEX)) + facet_grid(. ~SEX, space="free") + geom_histogram(bins = 14) + ggtitle("Histogram: Ratio Count by Sex")

#Boxplots
BoxPlotters <- ggplot(mydata, aes(y=RATIO, fill=SEX)) + geom_boxplot(color = "black",outlier.size=1.5, outlier.shape = .7, size = .25) + facet_grid(cols = vars(SEX)) + ggtitle("Boxplot: Ratio Count by Sex")
#Q-Q Plots
clusteredLine <- ggplot(mydata, aes(sample = RATIO, color= SEX)) + facet_grid(. ~SEX, space = "free") + geom_qq_line(color = "purple", cex=0.75) + geom_qq(cex=0.5, shape = 1)+ ggtitle("QQ: Ratio Count by Sex")

#Present in multiple lines.
grid.arrange(histogramSex, BoxPlotters, clusteredLine, nrow=3)

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (The histograms display that the counts for Female, infant and Male are heavily reflect the normal curve, with Female being the closest to an ideal normal distribution. )***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
#*boxplot.stats()* can be used to identify outlying values of a vector
femOutlier <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX =="F" ])$out, ]
InfantOutlier <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX =="I" ])$out, ]
MaleOutlier <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX =="M" ])$out, ]

#Display the observations by passing a data frame to the kable() function
kable(rbind(femOutlier,InfantOutlier, MaleOutlier))


```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (Most of the outlier ratios come from the infant abilenes, and most of those outliers are from the A1 Classification. For the female and male outliers, almost all the outliers are part of A3 classification with a few from A2. The majority of all outlier ratios are between .21 and .24 with a another at .3 and one at 0.06. The outlier ratios are pretty focused, but have a few outlier ratios themselves.)***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

#With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS
BoxVol<- ggplot(mydata, aes(x= CLASS, y=VOLUME, fill=CLASS)) + geom_boxplot(color = "black", size = .1, outlier.shape = 1) + scale_fill_brewer(palette = 1) + theme(legend.position = "none") + ggtitle("Class by Volume")
BoxWhole <- ggplot(mydata, aes(x = CLASS, y = WHOLE, fill = CLASS)) + geom_boxplot(color = "black", size = .1, outlier.shape = 1) + scale_fill_brewer(palette = 2) + theme(legend.position = "none") + ggtitle("Class by Whole Weight")

#Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS
ScatVol <- ggplot(mydata, aes(x= RINGS, y = VOLUME, fill = RINGS)) + geom_point(size = 0.75) + theme(legend.position = "none") + ggtitle("Rings by Volume")
ScatWhole<- ggplot(mydata, aes(x= RINGS, y = WHOLE, fill = RINGS)) + geom_point(size = 0.75) + theme(legend.position = "none") + ggtitle("Rings by Whole Weight")
 

#Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row.
grid.arrange(BoxVol, BoxWhole, ScatVol, ScatWhole, nrow=2)

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (Looking at volume and whole weight, the values seem to correlate to a wide range of rings and class values with overlap between the classes and corresponding volumnes and weights. For example a weight of 95 grams could correlate with A2 class or A3 class. or a volume of 500 could correlate to A4 and A5 classes. Would not be the most fool proof, but it would give an idea/range of age. But if we are looking to be specific and land in one class, volume and weight would not be the best predictors in my opinion. )***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
#Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS
AllAggreg <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN = mean, by = list(mydata$SEX, mydata$CLASS))

# Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS
MeanVol <- matrix(AllAggreg$VOLUME, nrow = 3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

MeanShuck <- matrix(AllAggreg$SHUCK, nrow = 3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

MeanRatio <- matrix(AllAggreg$RATIO, nrow = 3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

#Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose
kable(MeanVol, caption="Mean Volume by Class and Sex")
kable(MeanShuck, caption="Mean Shuck by Class and Sex")
kable(MeanRatio, caption="Mean Ratio by Class and Sex")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
#Present three graphs. Each graph should include three lines, one for each sex. 
colnames(AllAggreg) <- c('Sex', 'Class', 'Volume', 'Shuck', 'Ratio')

#The first should show mean RATIO versus CLASS
RatioVclass <- ggplot(data = AllAggreg, aes(x= Class, y = Ratio, group= Sex, color = Sex)) + geom_line(size = 0.5) + ggtitle("Class by Ratio")

#the second, mean VOLUME versus CLASS;
VolVclass <- ggplot(data = AllAggreg, aes(x= Class, y = Volume, group= Sex, color = Sex)) + geom_line(size = 0.5)+ ggtitle("Class by Volume")

#the third, mean SHUCK versus CLASS
ShuckVClass <- ggplot(data = AllAggreg, aes(x= Class, y = Shuck, group= Sex, color = Sex)) + geom_line(size = 0.5)+ ggtitle("Class by Shuck Weight")

#display all in one row
grid.arrange(RatioVclass, VolVclass, ShuckVClass, nrow=1)

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (I am extremely curious why weight and volume increases so rapidly and extremely from A1 to A4 then drops off rapidly (for weight) and stays somewhat even keel for volume. What would cause a significant reduction in weight but not volume? Shell density? Birth ability?  )***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
#The first line should show VOLUME by RINGS for the infants
InfantsBoxVolRing <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x= RINGS, y= VOLUME, fill = RINGS, group=RINGS)) + geom_boxplot(color = "red" , size = .1, outlier.shape = 1, fill = "black") + coord_cartesian(ylim= c(0,1000)) + ggtitle("Infants:Rings by Volume")

#Second for adults (both M and F) for volume by rings
AdultBoxVolRing <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX =="F")), aes(x= RINGS, y= VOLUME, fill = RINGS, group=RINGS)) + geom_boxplot(color = "red" , size = .1, outlier.shape = 1, fill = "blue")+ coord_cartesian(ylim= c(0,1000))+ ggtitle("Adults: Rings by Volume")

#The second line should show WHOLE by RINGS for the infants
InfantsBoxWholeRing <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x= RINGS, y= WHOLE, fill = RINGS, group=RINGS)) + geom_boxplot(color = "red" , size = .1, outlier.shape = 1, fill = "black")+ coord_cartesian(ylim= c(0,400))+ ggtitle("Infants: Rings by Whole Weight")

#Whole by Rings for Adults (M and F)
AdultBoxWholeRing <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX =="F")), aes(x= RINGS, y= WHOLE, fill = RINGS, group=RINGS)) + geom_boxplot(color = "red" , size = .1, outlier.shape = 1, fill = "blue")+ coord_cartesian(ylim= c(0,400))+ ggtitle("Adults: Rings by Whole Weight")

#present in grid
grid.arrange(InfantsBoxVolRing,AdultBoxVolRing, InfantsBoxWholeRing, AdultBoxWholeRing, nrow=2)
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (The growth of abalones happens rapidly from 1-11 rings, then evens out and slightly decreases from 12-15 rings for both volume and weight. Interestingly, infants have an extremely broad range of volume at 12 rings compared to adults. One thing I do find interesting is that infants can be considered infants all the way up to 14 rings, wouldn't they be adults by then? Did the study classify them as infants because they could not determine a gender? )***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (I believe the lack of direct correlation between volume and weight to age is part of why the study failed. I believe another element was the difficulty in determining sex for young abalone. There is a correlation between weight, shuck and volume for abalones related to their sex, and have a big bucket of infant abalone with not sex classification make it difficult to explore that correlation. Not knowing sex also makes it difficult to predict the breeding potential of the species, because you don't know how many females are able to reproduce. )***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (I would ask how does this sample reflect agains the total population? Is the sample as diverse in age, demographic, gender, socioeconomic status, etc as the main population? If the person doesn't know, I would ask for more data to do that analysis myself. I would also ask how long the data was collected for, as the data may be influenced by a catastrophic recent event so any results would be specific to that event and not to the population as a whole.  )***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Yes - observational studies are difficult as people are ultimately unreliable. How we interpret and document information is non-standardized so when determining causality, there are many factors that can influence the study and therefore the data. There are also many things we as humans learn to ignore, as our brains can only process so much. One of these elements we ignore could contribute to the causality of something and we wouldn't know because we didn't document it/collect info on it. However, observational studies are helpful for providing context around events and data sets, as well as giving you a direction to start looking when thinking of what to research next, or where data may be missing. In other words, observational studies are helpful with influence and understanding, but they very rarely give the whole picture and we should be cautious when relying on their results alone.  )***
